---
title: "class 18: Pertussis Resurgence Mini Project"
author: "Sawyer (PID: A16335277)"
format: pdf
---

First we will examine and explore Pertussis case numbers in the US as tracked by the CDC:
https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html


> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

We can use the datapasta package to scrape this data from the website into R:

```{r}
cdc <- data.frame(year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,2019L,2020L,2021L),
                  
cases = c(107473,164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116)
)

head(cdc)
```

I want a plot of cases per year with ggplot

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)

ggplot(cdc) + aes(year, cases) + geom_line() + geom_vline(xintercept = 1947, col='blue') +
  geom_vline(xintercept=1992, col='red') +
  geom_vline(xintercept=2020, col='orange')
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

About a couple decades after the introduction of the aP vaccine, we see a resurgence of pertussis cases (not as much as pre 1950, but still a considerable amount). A possible explanation could be that the newer generation of humans is resistant to the effects of the vaccine, making it less effective.


Access data from the CMI-PB project

This database (like many modern projects) uses an API to return JSON format data.

We will use the R package `jsonlite`.

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject)
```
> Q4. How many wP (the older whole-cell vaccine) individuals and aP (newer acellular vaccine) individuals are in this dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```


> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
head(subject$year_of_birth)
```

# Side-Note: Working with dates

We can use the lubridate package to ease the pain of doing math with dates.

```{r}
library(lubridate)
```

```{r}
today() - ymd("2000-01-01")
```

```{r}
today() - ymd("2002-6-25")
```

```{r}
time_length(today() - mdy("6-25-2002"), "years")
```

So what is the age of everyone on our dataset.

```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(dplyr)

#ap
ap <- subject %>% filter(infancy_vac == "aP")
round(summary(ap$age))
       
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round(summary(wp$age))

t.test(ap$age, wp$age)
```

> Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) + aes(age, fill=infancy_vac) + geom_histogram()
```


## Get more data from CMI-PB

```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = T)
head(specimen)
```

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

We need to **join** these two tables (subject and specimen) to make a single new "meta" table with all our metadata. We will use the `dyplyr` join functions to do this:

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Now we can read some of the other data from CMI-PB

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer",
                      simplifyVector = T)
head(ab_titer)
```

One more `inner_join()` to add all our metadata in `meta` on to our `ab_data` table:

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```


> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

2022 has the least amount of specimens collected in this dataset.


Our first exploratory plot:

```{r}
table(abdata$antigen)
```


```{r}
ggplot(abdata) + aes(MFI, antigen) + geom_boxplot()
```

Why are certain antigens and not others very variable in their detected levels here?

Can you facet or even just color by infancy_vac? Is there some difference?

```{r}
ggplot(abdata) + aes(MFI, antigen, color=infancy_vac) + geom_boxplot()
```

There are potentially some differences here but in general it is hard to tell with this whole dataset overview...

```{r}
table(abdata$dataset)
```

Let's focus in on just the 2021_dataset.

```{r}
abdata.21 <- filter(abdata, dataset == '2021_dataset')
table(abdata.21$dataset)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
igg <- abdata %>% filter(isotype == "IgG")

ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

PT and FIM2/3 show a wider range of normalized values. These were included in the vaccine, explaining why this might be the case.


> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


```{r}
filter(igg, antigen=="DT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


> Q16. What do you notice about these two antigens time courses and the PT data in particular?

PT and FIM2/3 appear to be generally increasing with the number of visits over time. This is does not appear to be true for DT. This could be indicative of increasing risk for Pertussis.



Focus on PT antigen for IgG levels
```{r}
pt.21 <- filter(abdata.21, isotype == 'IgG', antigen == 'PT')
```

plot of days (time) relative to boost vs MFI
```{r}
ggplot(pt.21) +
  aes(x=planned_day_relative_to_boost, y=MFI_normalised, col=infancy_vac, group=subject_id) + geom_point() + geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") + ylab("Normalized MFI") + xlab("planned day relative to boost")
```

> Q17. Do you see any clear difference in aP vs. wP responses?

In terms of the PT antigen, we see a clear increase in normalized MFI for the wP infancy vaccination subjects based on this line plot (at 14 days after the boost.
